## Allohexaploid *C. flexuosa*

### Introduction

*Cardamine flexuosa* (2*n* = 4*x* = 32, HHAA), wavy bittercress,
is an allopolyploid species that originated from two diploid progenitors,
*Cardamine hirsuta* (2*n* = 2*x* = 16, HH) and *Cardamine amara* (2*n* = 2*x* = 16, AA) [@ref_cflex_origin].
While *C. hirsuta* is typically adapted to dry habitats such as roadsides,
*C. amara* inhabits wet environments such as riversides or running streams.
The allopolyploid *C. flexuosa* thrives across a wide range of ecological habitats.

Akiyama et al. conducted an RNA-seq experiment using leaf samples of *C. flexuosa*
collected from wet and dry habitats across three different days [@ref_cflex_ecohabitats].
Each habitat–date combination included two or three biological replicates.
In the original study,
homeolog expression levels were quantified using read counts
processed through the HomeoRoq-sorting pipeline [@ref_homeoroq].
Analysis of homeolog expression ratios revealed that
a small percentage of homeologs in *C. flexuosa* exhibit shifts
in expression between wet and dry environments.

### Data Preparation

In this example, we demonstrate how to use HOBIT
to analyze this *C. flexuosa* dataset.
To reduce computation time, we focus on samples collected on May 16, 2013,
and randomly select 500 homeolog pairs (`c_flexuosa.0516.mini.txt.gz`).
Users can, however, analyze the full dataset or samples from other dates,
which are available in the [data](https://github.com/bitdessin/hespresso/tree/main/docs/data) directory,
and the same workflow can be applied.

```{r tt-cflex-load-counts}
gexp <- read.table("data/c_flexuosa.0516.mini.txt.gz",
                   header = TRUE, sep = "\t", row.names = 1)
head(gexp)
group <- c("wet", "wet", "wet", "dry", "dry", "dry")
```

Next, load the homeolog mapping table,
which defines the genes belonging to each homeolog pair.
In this tab-separated table, each row represents one homeolog pair,
and each column contains the corresponding gene from a subgenome.
In this example, the first and second columns contain gene names
from *C. hirsuta* and *C. amara*, respectively.

```{r tt-cflex-load-mappingtable}
mapping_table <- read.table("data/c_flexuosa.homeolog.txt.gz",
                            header = TRUE, sep = "\t")
head(mapping_table)
```

We then use the `newSeqCountData()` function
to organize the gene expression matrix (`gexp`) into a homeolog expression matrix
using the mapping table (`mapping_table`) and store the result in an `SeqCountData` class object.

```{r tt-cflex-create-expmx}
x <- newSeqCountData(gexp, group, mapping_table)
x
```

To visualize the distributions of homeolog expression ratios,
use the `plot_HER_distr()` function.
This function returns a list of histograms, one for each condition,
showing the distribution of homeolog expression ratios.

```{r tt-cflex-viz-distr}
distr_plots <- plot_HER_distr(x)
names(distr_plots)
```

By default, the expression ratio is calculated as the proportion contributed
by the first subgenome relative to all subgenomes.
In this dataset, the first subgenome is derived from *C. hirsuta*,
as indicated by the first column in the mapping table (`mapping_table`).

To display the distribution of homeolog expression ratios for each experimental group,
run the following code.

```{r tt-cflex-viz-distr-show, fig.cap="Distribution of homeolog expression ratios in _C. flexuosa_ under wet and dry habitats."}
library(ggplot2)
library(gridExtra)
grid.arrange(distr_plots[["wet"]] + ggtitle("wet"),
             distr_plots[["dry"]] + ggtitle("dry"),
             ncol = 2)
```

If no irregularities are observed after visually inspecting the distributions,
users can proceed with the statistical test.


### Statistical Test

HOBIT can be executed using the `hobit()` function, which performs a statistical test
to detect homeologs with shifts in expression ratios between wet and dry habitats.

```{r tt-cflex-hobit, results="hide", message=FALSE, warning=FALSE}
x_output <- hobit(x)
```

The order of the test output matches that of the input.

```{r tt-cflex-hobit-output}
head(x_output)
```

Note that the `pvalue` and `qvalue` columns in `x_output` may contain `NA` values.
This happens when at least one gene in a homeolog pair is not expressed in one or more conditions,
making it impossible to calculate expression ratios,
or when the full and/or reduced models cannot be constructed.
As a result, the statistical test cannot be performed for those pairs.

These `NA` values may affect sorting or counting the number of significant homeolog pairs.
Therefore, we recommend replacing `NA` with `1` (i.e., non-significant)
before performing downstream analyses.

```{r tt-cflex-hobit-replace-na}
x_output$pvalue[is.na(x_output$pvalue)] <- 1
x_output$qvalue[is.na(x_output$qvalue)] <- 1
```

To rank the output in descending order of *p*-values, use the following code.

```{r tt-cflex-hobit-sorted-output}
head(x_output[order(x_output$pvalue), ])
```

To visualize homeologs with shifts in expression ratios between wet and dry habitats,
use the `plot_HER()` function.
This function generates a scatter plot comparing the expression ratios
calculated for the wet and dry groups.
By default, the expression ratio is defined as the proportion contributed
by the first subgenome (i.e., the subgenome derived from *C. hirsuta*)
relative to the total expression across all subgenomes.
In the example below, homeologs with *q*-values less than 0.01 are highlighted.


```{r tt-cflex-viz-highlight, fig.cap="Changes in homeolog expression ratios between wet (x-axis) and dry (y-axis) habitats. Orange points indicate homeologs with significant shifts detected by HOBIT, while gray points represent homeologs without significant changes."}
is_sig <- ifelse(x_output$qvalue < 0.01, "q<0.01", "q>=0.01")
plot_HER(x, label = is_sig)
```

The plot generated by the `plot_HER()` function can also be visualized using `r pkg("plotly", "cran")`,
enabling interactive exploration of individual homeologs.
The interactive interface allows users to zoom in on specific regions of interest,
and hovering over data points displays detailed information for each homeolog.


```{r tt-cflex-viz-plotly, fig.cap="Visualization of homeolog expression ratio shifts with plotly."}
library(plotly)

p <- plot_HER(x, label = is_sig)
ggplotly(p)
```




### Analysis Environment

```{r tt-cflex-option-width, include=FALSE}
options(width = 120)
```

```{r tt-cflex-sessioninfo}
sessionInfo()
```
