## Allohexaploid *C. flexuosa*

```{r tt-cflex-setup, include=FALSE}
source("../_config.R")
```

### Introduction

*Cardamine flexuosa* (2*n* = 4*x* = 32, HHAA), wavy bittercress,
is an allopolyploid species that originated from two diploid progenitors,
*Cardamine hirsuta* (2*n* = 2*x* = 16, HH) and *Cardamine amara* (2*n* = 2*x* = 16, AA) [@ref_cflex_origin].
While *C. hirsuta* is typically adapted to dry habitats such as roadsides,
*C. amara* inhabits wet environments such as riversides or running streams.
The allopolyploid *C. flexuosa* thrives across a wide range of ecological habitats.

Akiyama et al. conducted an RNA-Seq experiment using leaf samples of *C. flexuosa*
collected from wet and dry habitats across three different days [@ref_cflex_ecohabitats].
Each habitat–date combination included two or three biological replicates.
Homeolog expression levels were quantified using read counts
processed through the HomeoRoq sorting pipeline [@ref_homeoroq].
Analysis of homeolog expression ratios revealed that
a small percentage of homeologs in *C. flexuosa* exhibit shifts
in expression depending on environmental differences, wet and dry.

### Data Preparation

In this example, we demonstrate how to use HOBIT and HomeoRoq to analyze the *C. flexuosa* dataset.
To reduce computation time, we focus on samples collected on May 16, 2013,
and randomly select 1,000 homeolog pairs (`c_flexuosa.0516.mini.txt.gz`).
Users can, however, analyze the full dataset or samples from other dates.
These datasets are available in the [data](https://github.com/bitdessin/hespresso/tree/main/docs/data) directory,
and the same workflow can be applied.

```{r tt-cflex-load-counts}
gexp <- read.table("../data/c_flexuosa.0516.mini.txt.gz", header = TRUE, sep = "\t", row.names = 1)
head(gexp)
group <- c("wet", "wet", "wet", "dry", "dry", "dry")
```

Next, load the homeolog mapping table,
which links gene expression values to their corresponding homeologs.
This table is a tab-separated file in which the first and second columns
represent gene names from *C. hirsuta* and *C. amara*, respectively.

```{r tt-cflex-load-mappingtable}
mapping_table <- read.table("../data/c_flexuosa.homeolog.txt.gz", header = TRUE, sep = "\t")
head(mapping_table)
```

We then use the `newExpMX()` function
to organize the gene expression matrix (`gexp`) into a homeolog expression matrix
using the mapping table (`mapping_table`) and store the result in an `ExpMX` class object.

```{r tt-cflex-create-expmx}
x <- newExpMX(gexp, group, mapping_table)
x
```

Before performing the test,
we normalize the raw read counts using the TMM method [@ref_tmm]
to adjust for differences in library size.
However, if the expression data (`gexp`) has already been normalized (e.g. FPKM),
this step can be skipped.

```{r tt-cflex-norm-counts}
x <- norm_counts(x)
```

To visualize the distributions of homeolog expression ratios,
use the `plot_HER_distr()` function.
This function returns a list of histograms, one for each condition,
showing the distribution of homeolog expression ratios.

```{r tt-cflex-viz-distr}
distr_plots <- plot_HER_distr(x)
names(distr_plots)
```

By default, the expression ratio is calculated as the proportion contributed
by the first subgenome relative to all subgenomes.
In this dataset, the first subgenome is derived from *C. hirsuta*,
as indicated by the first column in the mapping table (`mapping_table`).

To display the distribution of homeolog expression ratios for each experimental group,
run the following code.

```{r tt-cflex-viz-distr-show, fig.cap="Distribution of homeolog expression ratios in _C. flexuosa_ under wet and dry habitats."}
library(ggplot2)
library(gridExtra)

grid.arrange(distr_plots[["wet"]] + ggtitle("wet"),
             distr_plots[["dry"]] + ggtitle("dry"),
             ncol = 2)
```

After visually inspecting the distributions,
if no irregularities are observed, users can proceed to apply the statistical test.


### HOBIT

HOBIT can be executed using the `hobit()` function, which performs a statistical test
to detect homeologs with shifts in expression ratios between wet and dry habitats.
Analyzing 1,000 homeolog pairs with eight threads takes approximately five minutes.

```{r tt-cflex-hobit, results="hide", message=FALSE, warning=FALSE}
x_output <- hobit(x)
```

The order of the test output matches that of the input.

```{r tt-cflex-hobit-output}
head(x_output)
```

To rank the output in descending order of *p*-values, use the following code.

```{r tt-cflex-hobit-sorted-output}
head(x_output[order(x_output$pvalue), ])
```

To visualize homeologs with shifts in expression ratios between wet and dry habitats,
use the `plot_HER()` function.
This function generates a scatter plot comparing the expression ratios
calculated for the wet and dry groups.
By default, the expression ratio is defined as the proportion contributed
by the first subgenome (i.e., the subgenome derived from *C. hirsuta*)
relative to the total expression across all subgenomes.
In the example below, homeologs with *q*-values less than 0.01 are highlighted.


```{r tt-cflex-viz-highlight, fig.cap="Changes in homeolog expression ratios between wet (x-axis) and dry (y-axis) habitats. Orange points indicate homeologs with significant shifts detected by HOBIT, while gray points represent homeologs without significant changes."}
is_sig <- ifelse(x_output$qvalue < 0.01, "q<0.01", "q>=0.01")
plot_HER(x, label = is_sig)
```


### HomeoRoq

For allopolyploids composed of two subgenomes under two different conditions,
HomeoRoq is also a suitable option for detecting homeologs with shifts in expression ratios.
It can be executed using the `homeoroq()` function in the same way as `hobit()`,
as shown below.

```{r tt-cflex-hq-run, results="hide", message=FALSE, warning=FALSE}
x_homeoroq <- homeoroq(x)
```

```{r tt-cflex-hq-output}
head(x_homeoroq[order(x_homeoroq$pvalue), ])
```

The overlap of homeologs with shifts in expression ratios
detected by HOBIT and HomeoRoq can be visualized using an UpSet plot,
which is implemented in the `r pkg("UpSetR", "cran")` package [@ref_upsetr].


```{r tt-cflex-overlaps, fig.cap="Overlap of homeologs with shifted expression ratios detected by HOBIT and HomeoRoq."}
library("UpSetR")
sig_homeologs <- list(
    HOBIT = x_output$gene[x_output$qvalue < 0.01],
    HomeoRoq = x_homeoroq$gene[x_homeoroq$qvalue < 0.01]
)

upset(fromList(sig_homeologs))
```

Additionally, homeologs with shifts in expression ratios detected by both methods
can be visualized using the `plot_HER()` function.


```{r tt-cflex-overlaps-highlight, fig.cap="Comparison of homeologs with changes in expression ratios detected by HOBIT and HomeoRoq. Each point represents a homeolog expression ratio in wet (x-axis) and dry (y-axis) conditions. Points labeled as n.s. indicate homeologs without significant changes detected by either method; other points represent homeologs identified as having significant changes by the respective method."}
is_sig <- rep("n.s.", length = nrow(x_output))
is_sig[x_output$qvalue < 0.01] <- "HOBIT"
is_sig[x_homeoroq$qvalue < 0.01] <- "HomeoRoq"
is_sig[x_output$qvalue < 0.01 & x_homeoroq$qvalue < 0.01] <- "HOBIT&HomeoRoq"
is_sig <- factor(is_sig, levels = c("n.s.", "HOBIT", "HomeoRoq", "HOBIT&HomeoRoq"))
plot_HER(x, label = is_sig)
```


### Analysis Environment

```{r tt-cflex-sessioninfo}
sessionInfo()
```


