## Allotriploid *C. insueta*

### Introduction

Allotriploid *Cardamine insueta* (2n = 3x = 24, RRA) illustrates an uneven subgenome composition.
Two of its subgenomes are derived from *Cardamine rivularis* (2n = 2x = 16, RR),
a species inhabiting grassy areas away from riversides,
while the third subgenome comes from *Cardamine amara* (2n = 2x = 16, AA),
a species typically found on riversides or in streams [@ref_cins_origin_1; @ref_cins_origin_2].
RNA-seq studies on *C. insueta* leaflets floating on water suggested that
*C. insueta* shifts homeolog expression ratios over time following submergence [@ref_cins_submergestress].

### Data Preparation

Here, we demonstrate how to detect homeologs with changing expression ratios across time points using HOBIT,
similar to an ANOVA approach.
We use a sample dataset containing 500 homeolog pairs (`c_insueta.mini.txt.gz`) randomly selected from the original experiment,
although the full dataset is available in the [data](https://github.com/bitdessin/hespresso/tree/main/docs) directory
for users who wish to analyze it.
This dataset spans nine time points (0, 2, 4, 8, 12, 24, 48, 72, and 96 hours)
following leaf submergence and can be loaded using the following code.

```{r tt-cins-load-counts}
gexp <- read.table("data/c_insueta.mini.txt.gz",
                   header = TRUE, sep = "\t", row.names = 1)
head(gexp)
group <- c("T00", "T02", "T04", "T08", "T12", "T24", "T48", "T72", "T96")
```

Next, load the homeolog mapping table,
which defines the genes belonging to each homeolog tuple.
In this tab-separated table, each row represents one homeolog tuple,
and each column contains the corresponding gene from a subgenome.
In this example, the first and second columns contain gene names
from *C. rivularis* and *C. amara*, respectively.

```{r tt-cins-load-mappingtable}
mapping_table <- read.table("data/c_insueta.homeolog.txt.gz",
                            header = TRUE, sep = "\t")
head(mapping_table)
```

Use the `newSeqCountData()` function
to create a homeolog expression matrix from the gene expression data (`gexp`)
and mapping table (`mapping_table`).
The result is stored as an `SeqCountData` class object.

```{r tt-cins-create-expmx}
x <- newSeqCountData(gexp, group, mapping_table)
x
```

Users can visualize the distribution of homeolog expression ratios using the `plot_HER_distr()` function.
By default, the expression ratio is calculated as
the proportion contributed by the first subgenome relative to all subgenomes,
where the first subgenome corresponds to *C. rivularis*,
as listed in the first column of the mapping table (`mapping_table`).
The `plot_HER_distr()` function generates a histogram for each time point.

```{r tt-cins-viz-HER-distr, fig.cap="Distribution of homeolog expression ratios in _C. insueta_ across nine time points following leaf submergence in water."}
distr_plots <- plot_HER_distr(x)

library(ggplot2)
library(gridExtra)
distr_plots <- lapply(seq_along(distr_plots), function(p) {
    distr_plots[[p]] + ggtitle(names(distr_plots)[p])
})
grid.arrange(grobs = distr_plots, ncol = 3)
```

These visualizations indicate that most homeologs in *C. insueta* are expressed
at a 2:1 ratio between the *C. rivularis* and *C. amara* subgenomes,
as shown by histogram peaks centered around two-thirds.

### Statistical Test

Next, apply the `hobit()` function
to detect homeologs with differential expression ratios across the nine time points following leaf submergence.
Since this dataset contains only one replicate per condition,
set `no_replicate = TRUE` to avoid errors arising from dispersion estimation,
which requires multiple replicates.

```{r tt-cins-hobit, results="hide", message=FALSE, warning=FALSE}
x_output <- hobit(x, no_replicate = TRUE)
```

```{r tt-cins-hobit-output}
head(x_output)
```

Note that the `pvalue` and `qvalue` columns in `x_output` may contain `NA` values.
This happens when at least one gene in a homeolog pair is not expressed in one or more conditions,
making it impossible to calculate expression ratios,
or when the full and/or reduced models cannot be constructed.
As a result, the statistical test cannot be performed for those pairs.

These `NA` values may affect sorting or counting the number of significant homeolog pairs.
Therefore, we recommend replacing `NA` with `1` (i.e., non-significant)
before performing downstream analyses.
For example:


```{r tt-cins-hobit-replace-na}
x_output$pvalue[is.na(x_output$pvalue)] <- 1
x_output$qvalue[is.na(x_output$qvalue)] <- 1

head(x_output[order(x_output$pvalue), ])
```



### Analysis Environment

```{r tt-cins-option-width, include=FALSE}
options(width = 120)
```


```{r tt-cins-sessioninfo}
sessionInfo()
```
