## Allohexaploid *T. aestivum*

```{r tt-wheat-setup, include=FALSE}
source("../_config.R")
```

### Introduction

Bread wheat, *Triticum aestivum* (2n = 6x = 42, AABBDD),
is one of the world’s most important crops
and also serves as a model for studying allopolyploid species.
It originated through hybridization among three diploid progenitor species,
*Triticum urartu* (2n = 2x = 14, AA), 
an unknown species related to  *Aegilops speltoides* (2n = 2x = 14; SS,
which is thought to be the progenitor of BB),
and *Aegilops tauschii* (2n = 2x = 14, DD),
containing three distinct subgenomes [@ref_wheat_origin_1; @ref_wheat_origin_2; @ref_wheatgenome].
To demonstrate the applicability of HOBIT to allohexaploid species,
we present a case study using wheat RNA-seq data
to detect homeologs with shifts in expression ratios between two tissues,
shoot apex and leaf [@ref_wheatseedmatrix].


### Data Preparation

In this example, 500 homeolog triads were randomly selected from the original dataset.
The expression data can be loaded from the following file (`t_aestivum.w20.mini.txt.gz`).
Users may also choose to load the full dataset, which is stored
in the [data](https://github.com/bitdessin/hespresso/tree/main/docs/data) directory, for analysis.

```{r tt-wheat-load-counts}
gexp <- read.table("../data/t_aestivum.w20.mini.txt.gz",
                   header = TRUE, sep = "\t", row.names = 1)
head(gexp)
group <- c("apex", "apex", "apex", "leaf", "leaf", "leaf")
```

Next, load the mapping table that links each gene to its corresponding homeolog triads.
Since wheat has three subgenomes (i.e., A, B, and D),
the table contains three columns, with each column representing one subgenome.

```{r tt-wheat-load-mappingtable}
mapping_table <- read.table("../data/t_aestivum.homeolog.txt.gz",
                            header = TRUE, sep = "\t")
head(mapping_table)
```

We then construct an `ExpMX` class object using the `newExpMX()` function.
This converts the gene expression matrix (`gexp`)
into a homeolog expression matrix based on the mapping table (`mapping_table`).

```{r tt-wheat-create-expmx}
x <- newExpMX(gexp, group, mapping_table)
x
```

Before performing the test,
we normalize the raw read counts using the TMM method [@ref_tmm]
to adjust for differences in library size.
However, if the expression data (`gexp`) has already been normalized (e.g. FPKM),
this step can be skipped.

```{r tt-wheat-norm-counts}
x <- norm_counts(x)
```

To visualize the distributions of homeolog expression ratios, use the `plot_HER_distr()` function.
By default, the expression ratio is calculated as the proportion contributed
by the first subgenome relative to all subgenomes.
In this dataset, the first subgenome corresponds to the A-subgenome,
as indicated by the first column of the mapping table (`mapping_table`).


```{r tt-wheat-viz-distr, fig.cap="Distribution of homeolog expression ratios between shoot apex and leaf in the A-subgenome of wheat."}
distr_plots_A <- plot_HER_distr(x)
names(distr_plots_A)

library(ggplot2)
library(gridExtra)
grid.arrange(distr_plots_A[["apex"]] + ggtitle("A-ratio in apex sample"),
             distr_plots_A[["leaf"]] + ggtitle("A-ratio in leaf sample"),
             ncol = 2)
```

To calculate the ratios for the second and third subgenomes
(i.e., B-subgenome and D-subgenome), set `base = 2` and `base = 3`, respectively.


```{r tt-wheat-viz-distrs, fig.cap="Distribution of homeolog expression ratios between shoot apex and leaf across all subgenomes."}
distr_plots_B <- plot_HER_distr(x, base = 2)
distr_plots_D <- plot_HER_distr(x, base = 3)

grid.arrange(distr_plots_A[["apex"]] + ggtitle("A-ratio in apex sample"),
             distr_plots_A[["leaf"]] + ggtitle("A-ratio in leaf sample"),
             distr_plots_B[["apex"]] + ggtitle("B-ratio in apex sample"),
             distr_plots_B[["leaf"]] + ggtitle("B-ratio in leaf sample"),
             distr_plots_D[["apex"]] + ggtitle("D-ratio in apex sample"),
             distr_plots_D[["leaf"]] + ggtitle("D-ratio in leaf sample"),
             ncol = 2)
```

Histogram peaks at one-third for all subgenomes indicate
that most homeologs in wheat are equally expressed from the three subgenomes.

### Statistical Test

After visually inspecting the distributions of homeolog expression ratios,
use the `hobit()` function to identify homeologs
showing significant changes in expression ratios between shoot apex and leaf tissues.
This step takes approximately five minutes using eight threads and may vary depending on hardware performance.

```{r tt-wheat-run-hobit, results="hide", message=FALSE, warning=FALSE}
x_output <- hobit(x)
```

```{r tt-wheat-hobit-output}
head(x_output)
```

Note that the `pvalue` and `qvalue` columns in `x_output` may contain `NA` values.
This happens when at least one gene in a homeolog triad is not expressed in one or more conditions,
making it impossible to calculate expression ratios.
As a result, the statistical test cannot be performed for those triads.
These `NA` values may affect sorting or counting the number of significant homeolog triads.
Therefore, we recommend replacing `NA` with `1` (i.e., non-significant)
before performing downstream analyses.

```{r tt-wheat-hobit-replace-na}
x_output$pvalue[is.na(x_output$pvalue)] <- 1
x_output$qvalue[is.na(x_output$qvalue)] <- 1
```

The output can be sorted in ascending order of *p*-values using the following code.

```{r tt-wheat-hobit-sort-output}
head(x_output[order(x_output$pvalue), ])
```

Next, we visualize changes in expression ratios of the first subgenome between the two tissues.
In the plot, homeologs with *q*-values less than 0.01 are highlighted.

```{r tt-wheat-viz-HER-changes, fig.cap="Changes in homeolog expression ratios of the A-subgenome between shoot apex and leaf tissues. Each point represents the ratio in shoot apex (x-axis) and leaf (y-axis). Orange points indicate homeologs with significant changes, gray points indicate homeologs without significant changes."}
is_sig <- ifelse(x_output$qvalue < 0.01, "q<0.01", "q>=0.01")
plot_HER(x, label = is_sig)
```

To change the subgenome used for calculating expression ratios,
set the `base` argument to the corresponding subgenome number.
For example, to use the second subgenome, run the following.

```{r tt-wheat-viz-HER-changes-base-2, fig.cap="Changes in homeolog expression ratios of the B-subgenome between shoot apex and leaf tissues. Each point represents the ratio in shoot apex (x-axis) and leaf (y-axis). Orange points indicate homeologs with significant changes, gray points indicate homeologs without significant changes."}
plot_HER(x, base = 2, label = is_sig)
```


Note that expression ratios in the plots are calculated for the selected subgenome,
while highlighted homeologs are those that meet the significance threshold in any subgenome.
As a result, some homeologs may appear highlighted even if they show only minor changes in the plots,
because they may have significantly altered expression ratios in another subgenome.

### Combining Subgenomes

Users who want to test expression ratios
between the combined A- and B-subgenomes versus the D-subgenome
can first merge the expression of the A and B homeologs
using the `combine_hexp()` function.
After the merging step, statistical tests can be performed on the resulting dataset.

Assuming the A- and B-subgenomes are stored as the first and second subgenomes
in the `x` object (as defined in the `mapping_table`),
set `subgenomes = c(1, 2)` to combine them.


```{r tt-wheat-combine-subgenomes}
x_ABvsD <- combine_hexp(x, subgenomes = c(1, 2))
x_ABvsD
```

In this case, the combined A- and B-subgenomes are appended as the last subgenome,
so the subgenome order in `x_ABvsD` becomes D and A+B.
The distribution of homeolog expression ratios can
then be visualized using the `plot_HER_distr()` function.
By default, the ratios are calculated for the D-subgenome,
as D-subgenome is now the first subgenome in `x_ABvsD`.


```{r tt-wheat-viz-distr-combined, fig.cap="Distribution of homeolog expression ratios of D-subgenome between shoot apex and leaf in wheat."}
distr_plots_AB <- plot_HER_distr(x_ABvsD)
grid.arrange(distr_plots_AB[["apex"]] + ggtitle("AB-ratio in apex sample"),
             distr_plots_AB[["leaf"]] + ggtitle("AB-ratio in leaf sample"),
             ncol = 2)
```


The resulting distributions show a peak around one-thirds,
indicating that most homeologs are expressed equally between the D-subgenome
and the combined A- and B-subgenomes.

HOBIT can then be applied to the combined dataset as below.

```{r tt-wheat-combine-subgenomes-hobit, results="hide", message=FALSE, warning=FALSE}
x_ABvsD_output <- hobit(x_ABvsD)
x_ABvsD_output$pvalue[is.na(x_ABvsD_output$pvalue)] <- 1
x_ABvsD_output$qvalue[is.na(x_ABvsD_output$qvalue)] <- 1
```

```{r tt-wheat-combine-subgenomes-hobit-output}
head(x_ABvsD_output[order(x_ABvsD_output$pvalue), ])
```

The detection results can be visualized as follows.


```{r tt-wheat-viz-HER-changes-combined, fig.cap="Changes in homeolog expression ratios of the D-subgenome versus the combined A- and B-subgenomes between shoot apex and leaf tissues. Each point represents the ratio in shoot apex (x-axis) and leaf (y-axis). Orange points indicate homeologs with significant changes, gray points indicate homeologs without significant changes."}
is_sig <- ifelse(x_ABvsD_output$qvalue < 0.01, "q<0.01", "q>=0.01")
plot_HER(x_ABvsD, label = is_sig)
```



### Analysis Environment

```{r tt-wheat-option-width, include=FALSE}
options(width = 120)
```

```{r tt-wheat-sessioninfo}
sessionInfo()
```
