## Simulating Allotetraploid Data

The `r pkg("hespresso")` package provides functions
to simulate artificial read counts for allotetraploid and allohexaploid species.
By default, the simulation generates read counts
for an allotetraploid with 10,000 homeologs across two conditions,
each with three biological replicates.
The mean and dispersion parameters are estimated
from real RNA-seq data from an allotetraploid *Cardamine flexuosa* dataset [@ref_cflex_ecohabitats].

To generate a simulated dataset with default options, run as follows.

```{r sim-2x-default}
x <- sim_homeolog_counts()
x
```

Users can customize the simulation options.
For example, to simulate data with 1,000 homeologs and 5 replicates per condition,
run as follows.

```{r sim-2x-1000h-5r}
x <- sim_homeolog_counts(n_gene = 1000, n_replicates = c(5, 5))
x
```

The parameters used to simulate read counts are stored in the `x` object
returned by `sim_homeolog_counts()`.
Users can access these parameters through the `@meta` slot of the `x` object.
For example, `x@meta` contains the following information:

```{r sim-2x-meta-slotnames}
slotNames(x@meta)
```

To retrieve the ground-truth means used for the simulation,
users can access the `mu` slot as shown below.

```{r sim-2x-meta-mu}
names(x@meta@mu)
head(x@meta@mu[[1]])
head(x@meta@mu[[2]])
```

The dispersion parameters used in the simulation are estimated from
the relationship between mean and dispersion in a real RNA-seq dataset.
The fitted model function can be accessed via the `nls` slot.

```{r sim-2x-meta-nls}
x@meta@nls
```

The ground-truth labels for homeolog expression ratio (HER) shifts are stored in
`x@meta@is_shift`. Users can retrieve these labels directly from this slot or
call `def_sigShift(x)` without any additional arguments.

```{r sim-2x-meta-isshift}
table(x@meta@is_shift)
```


```{r sim-2x-meta-defsigshift-default}
gt <- def_sigShift(x)
table(gt)
```

The ground-truth HERs for each condition
can be retrieved directly from the `her` slot, as shown below.

```{r sim-2x-meta-her}
names(x@meta@her)
head(x@meta@her[[1]])
head(x@meta@her[[2]])
```

Using these ground-truth HERs,
users can also define ratio-shifted homeologs by setting thresholds for the
maximum absolute ratio difference (`Dmax`) and maximum odds ratio (`ORmax`).
These thresholds can be selected according to the true effect sizes of interest.
For example, the following code identifies homeologs with a `Dmax` greater than
0.2 or an `ORmax` exceeding 2.0.

```{r sim-2x-defgtrch}
is_sig <- def_sigShift(x, Dmax = 0.2, ORmax = 2, operator = "OR")
table(is_sig)
```

The output shows the number of homeologs meeting these thresholds.

To visualize the distribution of HERs across conditions,
use `plot_HER_distr()`.
This returns a list of histograms, one per condition,
with ratios calculated by default for the first subgenome (`base = 1`).

```{r sim-2x-plotherdistr}
distr_plots <- plot_HER_distr(x)
```

Since the simulated dataset contains two groups,
the `plot_HER_distr()` function returns two histograms.

```{r sim-2x-plotherdistr-plotnames}
names(distr_plots)
```

To display the distributions for each group, run the following code.

```{r sim-2x-plotherdistr-plots, fig.cap="Distribution of homeolog expression ratios in two simulated groups."}
library(ggplot2)
library(gridExtra)

grid.arrange(distr_plots[["group_1"]] + ggtitle("group_1"), 
             distr_plots[["group_2"]] + ggtitle("group_2"),
             ncol = 2)
```


To compare HERs between groups, use the `plot_HER()` function,
which by default plots HERs for the first subgenome in a scatter plot.


```{r sim-2x-plother, fig.cap="Changes in homeolog expression ratios between two groups. Each point represents the expression ratio in group 1 (x-axis) and group 2 (y-axis)."}
plot_HER(x, alpha = 0.3)
```

If the dataset was generated using `sim_homeolog_counts()`,
ground-truth significant shifts can be highlighted as follows using the `def_sigShift()` function.

```{r sim-2x-plother-label, fig.cap="Changes in homeolog expression ratios between two groups. Orange points indicate homeologs with significant changes in expression ratios, while gray points represent those without significant changes."}
is_sig <- def_sigShift(x)
plot_HER(x, label = is_sig, alpha = 0.3)
```
