## Simulating Allohexaploid Data

To simulate read counts for an allohexaploid with three subgenomes, set `n_subgenomes = 3`.
In this case, the mean and dispersion parameters are derived from real RNA-seq data
of allohexaploid wheat [@ref_wheatseedmatrix].

```{r sim-3x-default}
x <- sim_homeolog_counts(n_subgenomes = 3)
```

Simulation parameters can be retrieved from the `@meta` slot (i.e., `x@meta`).
The ground-truth labels for homeolog expression ratio (HER) shifts are stored in
`x@meta@is_shift`.
Users can retrieve these labels directly from this slot or call `def_sigShift(x)`.

```{r sim-3x-meta-isshift}
table(x@meta@is_shift)
```

```{r sim-3x-meta-defsigshift-default}
is_sig <- def_sigShift(x)
table(is_sig)
```

As in the allotetraploid example, users can specify `Dmax` and `ORmax` in
`def_sigShift()` to identify homeologs with significant HER shifts based on
the true effect-size thresholds of interest. Because an allohexaploid has
three subgenomes, specify which subgenome must satisfy the thresholds. For
example, to identify HER changes with `Dmax > 0.2` or `ORmax > 2.0` in the
first subgenome (`base = 1`), run:


```{r sim-3x-defsigshift-base1-params}
is_sig <- def_sigShift(x, base = 1, Dmax = 0.2, ORmax = 2.0, operator = "OR")
table(is_sig)
```

Visualization of allohexaploid data, comprising three subgenomes,
follows a similar workflow as for allotetraploids.
To display the distribution of HERs for the first subgenome,
run the following code.

```{r sim-3x-plotherdistr, fig.cap="Distribution of homeolog expression ratios for the first subgenome in a simulated allohexaploid across two groups."}
x <- sim_homeolog_counts(n_subgenomes = 3)
distr_plots_1 <- plot_HER_distr(x)
names(distr_plots_1)

library(ggplot2)
library(gridExtra)
grid.arrange(distr_plots_1[["group_1"]] + ggtitle("group_1 on 1st subgenome"), 
             distr_plots_1[["group_2"]] + ggtitle("group_2 on 1st subgenome"),
             ncol = 2)
```

Users can specify which subgenome to use for calculating HERs with the `base` argument.
For example, to visualize HERs for the second or third subgenome, run the following code.

```{r sim-3x-plotherdistr-base, fig.cap="Distribution of homeolog expression ratios for all subgenomes in a simulated allohexaploid across two groups."}
distr_plots_2 <- plot_HER_distr(x, base = 2)
distr_plots_3 <- plot_HER_distr(x, base = 3)

grid.arrange(distr_plots_1[["group_1"]] + ggtitle("group_1 on #1"),
             distr_plots_1[["group_2"]] + ggtitle("group_2 on #1"),
             distr_plots_2[["group_1"]] + ggtitle("group_1 on #2"),
             distr_plots_2[["group_2"]] + ggtitle("group_2 on #2"),
             distr_plots_3[["group_1"]] + ggtitle("group_1 on #3"),
             distr_plots_3[["group_2"]] + ggtitle("group_2 on #3"),
             ncol = 2)
```

The `plot_HER()` function also works for allohexaploids.
By default, it plots HERs for the first subgenome.

```{r sim-3x-plother, fig.cap="Changes in homeolog expression ratios of the first subgenome between two groups. Each point represents the expression ratio of the first subgenome in both groups."}
plot_HER(x, alpha = 0.3)
```

To plot HERs for a specific subgenome, set the `base` argument.
For example, to visualize the second subgenome, run the following code.

```{r sim-3x-plother-base2, fig.cap="Changes in homeolog expression ratios of the second subgenome between two groups. Each point represents the expression ratio of the second subgenome in both groups."}
plot_HER(x, base = 2, alpha = 0.3)
```

Ground-truth shifts can be highlighted in the scatter plot using `def_sigShift()`.
By default, shifts are defined across all subgenomes,
so some homeologs may appear highlighted even if changes in the plotted subgenome are minor.

```{r sim-3x-plother-gt, fig.cap="Changes in homeolog expression ratios of the first subgenome between two groups. Orange points indicate homeologs with significant changes in any subgenome, while gray points indicate no significant change."}
is_sig <- def_sigShift(x)
plot_HER(x, label = is_sig, alpha = 0.3)
```

To highlight only homeologs that meet the thresholds in a specific subgenome,
specify the `base` argument.

```{r sim-3x-plother-gt-base2, fig.cap="Changes in homeolog expression ratios of the first subgenome between two groups. Orange points indicate homeologs with significant changes in the first subgenome, while gray points show no significant change."}
is_sig_b1 <- def_sigShift(x, base = 1, Dmax = 0.2, ORmax = 2.0)
plot_HER(x, base = 1, label = is_sig_b1, alpha = 0.3)
```
