## Simulating Allotetraploid Data

```{r sim-2x-setup, include=FALSE}
source("../_config.R")
```

The `r pkg("hespresso")` package provides functions
to simulate artificial read counts for allotetraploid and allohexaploid species.
By default, the simulation generates read counts
for an allotetraploid with 10,000 homeologs across two conditions,
each with three biological replicates.
The mean and dispersion parameters are estimated
from real RNA-seq data from an allotetraploid *Cardamine flexuosa* dataset [@ref_cflex_ecohabitats].

To generate a simulated dataset with default options, run as follows.

```{r sim-2x-default}
x <- sim_homeolog_counts()
x
```

Options can be flexibly customized by users.
For example, to simulate data with 1,000 homeologs and 5 replicates per condition,
run as follows.

```{r sim-2x-1000h-5r}
x <- sim_homeolog_counts(n_gene = 1000, n_replicates = c(5, 5))
x
```

Parameters used for simulating read counts are stored in the `x` object
returned by the `sim_homeolog_counts()` function.
Users can access these parameters via the `@meta` slot of the `x` object.
For example, `x@meta` contains the following information:

```{r sim-2x-meta-slotnames}
slotNames(x@meta)
```

To retrieve the ground-truth means used for the simulation,
users can access the `mu` slot as follows.

```{r sim-2x-meta-mu}
names(x@meta@mu)
head(x@meta@mu[[1]])
head(x@meta@mu[[2]])
```

The dispersion parameters used in the simulation are estimated from
the relationship between mean and dispersion in a real RNA-seq dataset.
The fitted model function can be accessed via the `nls` slot.

```{r sim-2x-meta-nls}
x@meta@nls
```

Additionally, the ground-truth homeolog expression ratios can be calculated manually
from the ground-truth means, or they can be retrieved directly from the `her` slot, as follows.

```{r sim-2x-meta-her}
names(x@meta@her)
head(x@meta@her[[1]])
head(x@meta@her[[2]])
```

Using this information,
users can define homeologs that show significant changes in expression ratios between conditions.
Alternatively, the `def_sigShift()` function can automatically identify homeologs
with significant shifts without the need for manual calculations.
For example, to identify homeologs with a maximum absolute ratio difference (`Dmax`) greater than 0.2
or a maximum odds ratio (`ORmax`) exceeding 2.0, use the following code.

```{r sim-2x-defgtrch}
x <- sim_homeolog_counts()

is_sig <- def_sigShift(x, Dmax = 0.2, ORmax = 2, operator = "OR")
table(is_sig)
```

The output shows the number of homeologs meeting these thresholds.

To visualize the distribution of expression ratios across conditions,
use the `plot_HER_distr()` function.
This returns a list of histograms, one per condition,
with ratios calculated by default for the first subgenome (`base = 1`).

```{r sim-2x-plotherdistr}
x <- sim_homeolog_counts()
distr_plots <- plot_HER_distr(x)
```

Since the simulated dataset contains two groups,
the `plot_HER_distr()` function returns two histograms.

```{r sim-2x-plotherdistr-plotnames}
names(distr_plots)
```

To display the distributions for each group, run as follows.

```{r sim-2x-plotherdistr-plots, fig.cap="Distribution of homeolog expression ratios in two simulated groups."}
library(ggplot2)
library(gridExtra)

grid.arrange(distr_plots[["group_1"]] + ggtitle("group_1"), 
             distr_plots[["group_2"]] + ggtitle("group_2"),
             ncol = 2)
```


To compare expression ratios between groups, use the `plot_HER()` function,
which by default plots ratios for the first subgenome in a scatter plot.


```{r sim-2x-plother, fig.cap="Changes in homeolog expression ratios between two groups. Each point represents the expression ratio in group 1 (x-axis) and group 2 (y-axis)."}
plot_HER(x, alpha = 0.3)
```

If the dataset was generated using `sim_homeolog_counts()`,
ground-truth significant shifts can be highlighted as follows using the `def_sigShift()` function.

```{r sim-2x-plother-label, fig.cap="Changes in homeolog expression ratios between two groups. Orange points indicate homeologs with significant changes in expression ratios, while gray points represent those without significant changes."}
is_sig <- def_sigShift(x, Dmax = 0.2, ORmax = 2.0)
plot_HER(x, label = is_sig, alpha = 0.3)
```

