## Simulation Based Evaluation

### Standard Protocol

The combination of `sim_homeolog_counts()` and `def_sigShift()`
allows users to evaluate HOBIT's performance using standard metrics,
such as precision, recall, F1 score, and area under the ROC curve (AUC).
Here, we demonstrate this workflow with a small simulated read-count dataset.

First, define a function to calculate metrics from the test results.
The function identifies predicted homeologs with significant expression-ratio shifts
using a false discovery rate (FDR) threshold of 0.05.
It then compares these predictions with the ground-truth shifts defined by
`def_sigShift()`.
The AUC is calculated from the *p*-values in the `hobit()` output
using the `auc()` and `roc()` functions from the `r pkg("pROC", "cran")` package [@ref_proc].


```{r sim-fw-def-calc-metrics}
library(pROC)

calc_metrics <- function(output, q_cutoff = 0.05) {
    # Replace NA values with 1.0 (non-significant).
    output$pvalue[is.na(output$pvalue)] <- 1
    output$qvalue[is.na(output$qvalue)] <- 1

    # Define the ground truth and predictions.
    gt <- def_sigShift(x)
    pred <- (output$qvalue < q_cutoff)

    # Calculate performance metrics.
    tp <- sum(pred & gt)
    tn <- sum(!pred & !gt)
    fp <- sum(pred & !gt)
    fn <- sum(!pred & gt)

    precision <- tp / (tp + fp)
    recall <- tp / (tp + fn)
    f1 <- 2 * precision * recall / (precision + recall)
    auc <- auc(roc(gt, 1 - output$pvalue, levels = c(FALSE, TRUE), direction = "<"))

    c(precision = precision, recall = recall, f1 = f1, auc = auc)
}
```

Then, generate a set of simulated read counts.

```{r sim-fw-dataset, results="hide", message=FALSE, warning=FALSE}
x <- sim_homeolog_counts(n_genes = 1000)
```

Next, run HOBIT MLE and use `calc_metrics()` to evaluate its performance.


```{r sim-fw-hobit-mle-metrics, message=FALSE, warning=FALSE}
x_output_mle <- hobit(x)
calc_metrics(x_output_mle)
```

The same procedure can be applied to the other methods.

```{r sim-fw-methods-metrics, eval=FALSE, message=FALSE, warning=FALSE}
x_output_mcmc_strict <- hobit(x, method = "mcmc", mode = "strict")
calc_metrics(x_output_mcmc_strict)

x_output_mcmc_fast <- hobit(x, method = "mcmc", mode = "fast")
calc_metrics(x_output_mcmc_fast)

x_output_homeoroq <- homeoroq(x)
calc_metrics(x_output_homeoroq)
```


To reduce computation time, this example simulates only 1,000 homeologs.
For reliable evaluation, however, we recommend simulating a larger number
of homeologs so that the dataset contains enough significant expression-ratio shifts
for robust validation.


### Large-scale Simulation Evaluation

Using this standard protocol, we generated and evaluated simulated datasets
10 times for each combination of `n_subgenomes` (`2` and `3`) and `n_reps` (`3` and `5`).
The complete evaluation script is available at
[docs/simdata_validation_framework.R](https://github.com/bitdessin/hespresso/blob/main/docs/simdata_validation_framework.R)
in the `r pkg("hespresso")` repository.
HomeoRoq was not evaluated when `n_subgenomes = 3`
because it supports only allopolyploids with two subgenomes.

The full evaluation takes approximately three days on an Ubuntu system with
four threads and an Intel(R) Core(TM) i9-11900K processor running at 3.50 GHz.
The resulting performance metrics are summarized in the figure below.

```{r sim-fw-full-evaluation-metrics-viz, echo = FALSE, fig.height = 7.4,  fig.cap="Evaluation metrics. Simulation conditions are represented by the subplot columns, and metrics by the rows. In each subplot, the x-axis shows the method: MLE, MCMC (strict), or MCMC (fast). The y-axis shows the corresponding metric. Each point represents one simulation result."}
library(dplyr)
library(tidyr)
library(ggplot2)
x <- read.table("data/hobit_metrics.tsv.gz", header = TRUE, sep = '\t') %>%
    mutate(model = factor(model, levels = c("MLE", "MCMC (strict)", "MCMC (fast)", "HOMEOROQ")))

x %>%
    pivot_longer(cols = c(precision, recall, f1, auc), names_to = 'metric', values_to = 'value') %>%
    mutate(metric = factor(metric, levels = c('auc', 'f1', 'precision', 'recall'))) %>%
    ggplot(aes(x = model, y = value, group = model)) +
        geom_violin() +
        geom_jitter(width = 0.2, alpha = 0.5) +
        facet_grid(metric ~ n_subgenomes + n_replicates, scales = 'free_y',
                   labeller = labeller(.rows = label_value, .cols = label_both)) +
        theme(axis.text.x = element_text(angle = 30, hjust = 0.5))
```

The execution-time results show that HOBIT MLE is the fastest method,
completing in a few minutes. HOBIT MCMC strict is the slowest method,
taking approximately 60 minutes, depending on the number of subgenomes and
replicates. HOBIT MCMC fast takes approximately half as long as MCMC strict.

```{r sim-fw-full-evaluation-exetime-viz, echo = FALSE, fig.height = 3, fig.cap="Execution time (min). Simulation conditions are represented by the subplot columns. In each subplot, the x-axis shows the method: MLE, MCMC (strict), or MCMC (fast). Each point represents one simulation result."}
x %>%
    mutate(exe_time = exe_time / 60) %>%
    ggplot(aes(x = model, y = exe_time, group = model)) +
        geom_violin() +
        geom_jitter(width = 0.2, alpha = 0.5) +
        facet_grid(~ n_subgenomes + n_replicates, scales = 'free_y',
                labeller = labeller(.rows = label_value, .cols = label_both)) +
        theme(axis.text.x = element_text(angle = 30, hjust = 0.5)) +
        ylab('execution time [min]')
```
