## HOBIT

### Model

HOBIT [@ref_hobit] is a statistical method for detecting
ratio-shifted homeologs (RSHs), whose homeolog expression ratios (HERs)
differ among experimental conditions. For each homeolog tuple, HOBIT compares
a full model that allows the HER to vary among conditions with a reduced model
that assumes a common HER across conditions.

Because RNA-seq read counts commonly exhibit overdispersion, HOBIT
models homeolog expression using a negative binomial (NB) distribution:

$$
f_{\mathrm{NB}}(x \mid \mu, \phi) =
\binom{x+\phi-1}{x}
\left(\frac{\mu}{\mu+\phi}\right)^x
\left(\frac{\phi}{\mu+\phi}\right)^\phi,
$$

where $\mu$ and $\phi$ are the mean and dispersion parameters, respectively.
Under this parameterization,

$$
\operatorname{E}(X)=\mu, \qquad
\operatorname{Var}(X)=\mu+\frac{\mu^2}{\phi}.
$$

Let $x_{ik}^{c}$ denote the normalized expression count for subgenome $k$ in
biological replicate $i$ under condition $c$. For example, in an experiment
with three biological replicates of allohexaploid wheat under control ($C$)
and treatment ($T$) conditions, $k \in \{A,B,D\}$, $i \in \{1,2,3\}$, and
$c \in \{C,T\}$.

For condition $c$, let $\mu^{c}$ denote the mean total expression of the
homeolog tuple across all subgenomes, let $\phi^{c}$ denote its dispersion,
and let

$$
\boldsymbol{\theta}^{c}
=
\left(\theta_{1}^{c},\ldots,\theta_{K}^{c}\right)
$$

be the HER vector, subject to

$$
\theta_{k}^{c} \geq 0, \qquad
\sum_{k=1}^{K}\theta_{k}^{c}=1.
$$

The likelihood of the observed homeolog counts under condition $c$ is

$$
\mathcal{L}
\left(
\boldsymbol{\theta}^{c},
\mu^{c},
\phi^{c}
\right)
=
\prod_{k}\prod_{i}
f_{\mathrm{NB}}
\left(
x_{ik}^{c}
\mid
\theta_{k}^{c}\mu^{c},
\phi^{c}
\right).
$$

In the original HOBIT algorithm, $\mu^{c}$ and
$\boldsymbol{\theta}^{c}$ are estimated by Markov chain Monte Carlo (MCMC)
sampling from non-informative prior distributions. The dispersion
$\phi^{c}$ is estimated in advance using `estimateDisp()` from the
`r pkg("edgeR", "bioc")` package [@ref_edger], which shares information across
homeologs and stabilizes dispersion estimation.

To test whether HERs differ among conditions, HOBIT compares the full and
reduced models. For the two-condition example, the full model allows separate
HER vectors under the control and treatment conditions:

$$
\log \mathcal{L}_{\mathrm{full}}
=
\sum_{c=C,T}
\log \mathcal{L}
\left(
\boldsymbol{\theta}^{c},
\mu^{c},
\phi^{c}
\right).
$$

The reduced model represents the null hypothesis and uses a shared HER vector,
$\boldsymbol{\theta}^{S}$, under both conditions:

$$
\log \mathcal{L}_{\mathrm{reduced}}
=
\sum_{c=C,T}
\log \mathcal{L}
\left(
\boldsymbol{\theta}^{S},
\mu^{c},
\phi^{c}
\right),
$$

where the original algorithm defines

$$
\boldsymbol{\theta}^{S}
=
\frac{
\boldsymbol{\theta}^{C}
+
\boldsymbol{\theta}^{T}
}{2}.
$$

For more than two conditions, the corresponding shared vector is obtained from
the condition-specific HER vectors, and the likelihood is summed over all
conditions. Under the null hypothesis,

$$
\boldsymbol{\theta}^{C}
=
\boldsymbol{\theta}^{T}
=
\boldsymbol{\theta}^{S},
$$

so the full and reduced models have the same likelihood in theory.

After calculating the two likelihoods, the original HOBIT algorithm performs a
likelihood ratio test. The unnormalized p-value is

$$
P_{\mathrm{raw}}
=
\chi^{2}
\left(
-2
\left(
\log \mathcal{L}_{\mathrm{reduced}}
-
\log \mathcal{L}_{\mathrm{full}}
\right),
(C-1)(K-1)
\right),
$$

where $C$ and $K$ are the numbers of conditions and subgenomes, respectively,
and $\chi^{2}(t,\nu)$ denotes the upper-tail probability of a chi-squared
variable with $\nu$ degrees of freedom evaluated at $t$.


### Implementation

The current `r pkg("hespresso")` package implements three variants of HOBIT:

- HOBIT MLE
- HOBIT MCMC fast
- HOBIT MCMC strict


**HOBIT MLE** {#hobit-mle}

HOBIT MLE optimizes the full and reduced models independently and calculates
*p*-values from the resulting maximum-likelihood estimates (MLE).
The likelihood-ratio statistic is

$$
2
\left(
\log \mathcal{L}_{\mathrm{full}}
-
\log \mathcal{L}_{\mathrm{reduced}}
\right).
$$

HOBIT MLE requires raw count data and incorporates library size normalization internally.
Without MCMC sampling, the analysis of
10,000 homeolog tuples takes several minutes.


**HOBIT MCMC strict** {#hobit-mcmc-strict}

HOBIT MCMC in strict mode uses a stricter model comparison procedure by sampling
the full and reduced models independently. Although it requires nearly twice
the execution time of fast mode---approximately 40 minutes for 10,000 homeolog
tuples---it provides an explicit fit for each hypothesis.

This variant incorporates library size normalization internally by applying
sample-specific normalization offsets to raw count data. Likelihood summaries
are based on posterior samples rather than MLE.


**HOBIT MCMC fast** {#hobit-mcmc-fast}

HOBIT MCMC in fast mode constructs the full model by MCMC sampling and
constructs the reduced model using a shared HER derived from the full model
posterior. Because the reduced model is not sampled independently, this mode
reduces the execution time by approximately half, requiring about 20 minutes
for 10,000 homeolog tuples. Likelihood summaries are based on posterior samples
rather than MLE, and *p*-values are computed from the posterior samples.

Because the likelihood difference can become increasingly sensitive as the
number of biological replicates increases, HOBIT MCMC fast additionally
applies an ad hoc normalization using the average number of replicates per
condition, $R$, and reports the resulting value as the p-value:

$$
P
=
\chi^{2}
\left(
-2
\left(
\log \mathcal{L}_{\mathrm{reduced}}
-
\log \mathcal{L}_{\mathrm{full}}
\right)
\left(
\frac{\log R+1}{R}
\right),
(C-1)(K-1)
\right).
$$

This variant (i.e., HOBIT MCMC fast) corresponds to the implementation
evaluated and described in the original paper [@ref_hobit].


### Inputs

The HOBIT functions require an input object of class `SeqCountData`.
To create this object from real RNA-seq data,
users need to provide the following three components:

- a gene expression matrix (i.e., raw read counts)
- a vector or data frame of condition labels
- a mapping table to organize gene expression data into homeolog expression

The following example shows how to load gene expression data,
define experimental conditions, and load the mapping table from a file.

The gene expression matrix should be a data frame or matrix,
with row names corresponding to gene names.

```{r fc-hobit-input-preparation-gexp}
gexp <- read.table("data/c_flexuosa.0516.mini.txt.gz",
                   header = TRUE, sep = "\t", row.names = 1)
head(gexp)
```

Note that all HOBIT variants require raw, unnormalized read counts.
Do not provide transcripts per million (TPM),
fragments per kilobase of transcript per million mapped reads (FPKM),
counts per million (CPM), normalized counts, or log-transformed expression values.


The condition labels can be provided as a vector or a data frame.

```{r fc-hobit-input-preparation-mappingtable}
group <- c("wet", "wet", "wet", "dry", "dry", "dry")
group
```

The mapping table (`mapping_table`) is a data frame
in which the first and second columns correspond to gene names
from _C. hirsuta_ and _C. amara_, respectively.
It is assumed that all gene names in both columns appear
in the row names of the gene expression matrix (`gexp`).

```{r fc-hobit-input-preparation-group}
mapping_table <- read.table("data/c_flexuosa.homeolog.txt.gz",
                            header = TRUE, sep = "\t")
head(mapping_table)
```

Once the three required components are prepared,
an `SeqCountData` object can be created using the `newSeqCountData()` function.

```{r fc-hobit-input-newSeqCountData}
x <- newSeqCountData(gexp, group, mapping_table)
x
```


### Test


**HOBIT MLE**

HOBIT MLE can be run with the `hobit()` function using its default settings,
in which `method = "mle"` is set.

```{r fc-hobit-run-mle, results="hide", message=FALSE, warning=FALSE}
x_output_mle <- hobit(x)
```

Alternatively, users can specify the `method` argument explicitly.

```{r fc-hobit-run-mle-2, eval=FALSE, message=FALSE, warning=FALSE}
x_output_mle <- hobit(x, method = "mle")
```


**HOBIT MCMC strict**

HOBIT MCMC strict mode can be run with the `hobit()` function using
`method = "mcmc"`.


```{r fc-hobit-run-mcmc-strict, eval=FALSE, message=FALSE, warning=FALSE}
x_output_strict <- hobit(x, method = "mcmc")
```

Alternatively, the following usage has the same effect.


```{r fc-hobit-run-mcmc-stric-2, eval=FALSE, message=FALSE, warning=FALSE}
x_output_strict <- hobit(x, method = "mcmc", mode = "strict")
```


**HOBIT MCMC fast**


HOBIT MCMC fast mode should be run with the corresponding arguments
of the `hobit()` function.

```{r fc-hobit-run-mcmc-fast, eval=FALSE, message=FALSE, warning=FALSE}
x_output_fast <- hobit(x, method = "mcmc", mode = "fast")
```


### Outputs

Each method returns a data frame with one row per homeolog tuple, and the row
order matches the input. All three variants report the following principal
columns:

- `gene`: homeolog-tuple identifier
- `pvalue`: p-value from the likelihood ratio test
- `qvalue`: Benjamini--Hochberg-adjusted p-value
- `theta0__*`: estimated HER shared across conditions under the reduced model
- `theta1__*__*`: condition-specific HER estimates under the full model
- `D__*`: differences in individual subgenome proportions
- `logLik_H0` and `logLik_H1`: log-likelihood summaries for the reduced and full models
- `lrt`: likelihood-ratio statistic
- `df`: asymptotic degrees of freedom
- `LR__*`: changes in pairwise subgenome log-ratios
- `RR__*`: exponentiated pairwise log-ratio changes
- `Dmax`, `LRmax`, and `RRmax`: maximum effect-size summaries
- `status`: fitting and testing status


```{r fc-hobit-output-print}
head(x_output_mle)
```

The `pvalue` and `qvalue` columns may contain `NA` values when a test cannot be
completed, for example because a condition has zero total expression or because
model fitting fails. The `status` column returned by the MLE and strict MCMC
implementations should be inspected before downstream analysis.

If failed tests are to be treated as non-significant in an operational analysis,
replace missing values with `1` explicitly:

```{r fc-hobit-output-print-na}
x_output_mle$pvalue[is.na(x_output_mle$pvalue)] <- 1
x_output_mle$qvalue[is.na(x_output_mle$qvalue)] <- 1
```


### Options

#### Options for HOBIT MLE

##### MLE parallelization

The `n_threads` argument controls parallel processing across homeolog tuples.

```{r fc-hobit-mle-nthreads, eval=FALSE}
output_mle <- hobit(x, method = "mle", n_threads = 8)
```


##### MLE optimization options

The `gradient_tolerance` argument specifies the maximum absolute gradient
allowed after optimization. Fits exceeding this threshold are reported as
failed by the current implementation. The default is `1e-2`.
The output columns `max_gradient_H0` and `max_gradient_H1` can be inspected when
many tuples return `h0_fit_failed` or `h1_fit_failed`.

```{r fc-hobit-mle-gradient, eval=FALSE}
output_mle <- hobit(x, method = "mle", gradient_tolerance = 0.1)
```

Relaxing the gradient threshold may recover additional fits,
but its effects on fitting coverage, type-I error, and LRT calibration should be
checked before routinely using a less stringent value.

The `lrt_tolerance` argument controls how slightly negative LRT values caused by
numerical error are handled. Values between `-lrt_tolerance` and zero are set to
zero; more negative values are reported as `lrt_failed`.

Named arguments supplied through `...` are added to the `control` list of
`nlminb()`. Common options include `iter.max`, `eval.max`, `rel.tol`, `x.tol`,
and `trace`.

```{r fc-hobit-mle-nlminb, eval=FALSE}
output_mle <- hobit(x, method = "mle",
    iter.max = 5000, eval.max = 5000, rel.tol = 1e-12, x.tol = 1e-10)
```

When optimization failures occur, first inspect `status`, the optimizer return
codes, and the maximum-gradient columns. Increasing iteration limits is useful
only when the optimizer has not already terminated successfully; a return code
of zero accompanied by a gradient slightly above the acceptance threshold may
instead indicate that the gradient criterion is stricter than the optimizer's
own convergence criterion.


##### Other options

The `no_replicate` argument selects the approximation used when biological
replicates are unavailable. With replicated data, edgeR estimates a common
tagwise dispersion for each homeolog using all samples and a design matrix
containing the experimental group; the estimated value is then shared across
conditions during model fitting.

```{r fc-hobit-mle-norep, eval=FALSE}
output_mle <- hobit(x, method = "mle", no_replicate = TRUE)
```

Inference based on unreplicated data should therefore be interpreted cautiously.


#### Options for HOBIT MCMC

##### Stan parallel processing

The `n_threads` argument controls parallel processing across homeolog tuples.
By default, it follows the global `mc.cores` option, or uses one worker if that
option is not set.

```{r fc-hobit-mcmc-nthreads, eval=FALSE}
output_strict <- hobit(x, method = "mcmc", mode = "strict",
    n_threads = 8)
```

The `parallel_chains` argument controls the number of Stan chains run in
parallel for each model and is passed to `cmdstanr`.

```{r fc-hobit-mcmc-parallel-chains, eval=FALSE}
output_strict <- hobit(x, method = "mcmc", mode = "strict",
    n_threads = 1, chains = 4, parallel_chains = 4)
```

`n_threads` and `parallel_chains` act at different levels.
If both exceed one, the analysis may use as many as
`n_threads` $\times$ `parallel_chains` concurrent processes.


##### Stan sampling options

Additional arguments in `...` are passed to the `sample()` method of
`cmdstanr`. Important options include:

- `chains`: number of MCMC chains (default in `cmdstanr`: `4`)
- `iter_warmup`: number of warmup iterations per chain
- `iter_sampling`: number of post-warmup iterations per chain
- `save_warmup`: whether warmup draws are saved
- `thin`: interval at which post-warmup draws are retained
- `max_treedepth`: maximum NUTS tree depth
- `adapt_delta`: target acceptance probability
- `seed`: random-number seed

```{r fc-hobit-mcmc-iters, eval=FALSE}
output_strict <- hobit(x, method = "mcmc", mode = "strict",
    chains = 4, iter_warmup = 2000, iter_sampling = 5000)
```

If divergent transitions or tree-depth warnings occur,
increasing `adapt_delta` or `max_treedepth` may improve sampling,
at the cost of additional computation.

```{r fc-hobit-mcmc-depth, eval=FALSE}
output_strict <- hobit(x, method = "mcmc", mode = "strict",
    adapt_delta = 0.95, max_treedepth = 12)
```


##### Stan verbosity options

Console output from Stan can be controlled with options such as:

- `refresh`: frequency of sampling progress updates; set to `0` to suppress them
- `show_messages`: whether Stan diagnostic and progress messages are printed

```{r fc-hobit-mcmc-verbosity, eval=FALSE}
output_strict <- hobit(x, method = "mcmc", mode = "strict",
    refresh = 0, show_messages = FALSE)
```

For additional options, refer to the
[cmdstanr sampling documentation](https://mc-stan.org/cmdstanr/reference/model-method-sample.html).


##### Other options

The `no_replicate` argument selects the approximation used when biological
replicates are unavailable.

```{r fc-hobit-mcmc-norep, eval=FALSE}
output_strict <- hobit(x, method = "mcmc", mode = "strict", no_replicate = TRUE)
```

However, dispersion cannot be estimated reliably
from an experiment with no biological replication.
Therefore, `no_replicate = TRUE` should be regarded as an
approximation rather than a substitute for a replicated experimental design.
