## HomeoRoq

```{r fc-hq-setup, include=FALSE}
source("../_config.R")
```

HomeoRoq [@ref_homeoroq] estimates the probability that
homeolog expression ratios remain constant
between two experimental groups using RNA-Seq read count data for each homeolog.
Since this probability cannot be derived analytically,
HomeoRoq adopts a Bayesian framework
to approximate the posterior distribution of homeolog expression ratios
under the null hypothesis of no change between conditions.
The `homeoroq()` function provides a complete interface
for detecting homeologs whose expression ratios differ between two conditions.

By default, it performs 10,000 sampling iterations (`iter_sampling = 1e4`) per probability calculation
and repeats the calculation four times (`chians = 4`).
However, for robust and reproducible estimation of statistical significance (i.e., *p*-values),
it is recommended to run multiple independent sampling replicates,
each with a sufficiently large number of iterations.
To reproduce the settings used in the original HomeoRoq implementation [@ref_homeoroq],
set `iter_sampling = 1e4` and `chains = 10`.


### Inputs

The `homeoroq()` function requires an input object of class `ExpMX`,
same to the `hobit()` function.


```{r fc-hq-input-newExpMX}
gexp <- read.table("../data/c_flexuosa.0516.mini.txt.gz", header = TRUE, sep = "\t", row.names = 1)
group <- c("wet", "wet", "wet", "dry", "dry", "dry")
mapping_table <- read.table("../data/c_flexuosa.homeolog.txt.gz", header = TRUE, sep = "\t")
x <- newExpMX(gexp, group, mapping_table)
```

### Test

Once an `ExpMX` object is prepared,
normalize the counts if the data have not already been normalized,
then run the test using the `homeoroq()` function.

```{r fc-hq-run-homeoroq, results="hide", message=FALSE, warning=FALSE}
x <- norm_counts(x)
x_output <- homeoroq(x)
```


### Outputs

The result (`x_output`) is a `data.frame` class object with one row per homeolog,
summarizing the statistical test results:


- `gene`: Homeolog ID. The gene names defined in the first column of the mapping table.
- `pvalue`: *p*-value from the likelihood ratio test.
- `qvalue`: Adjusted *p*-value using the Benjamini-Hochberg method.
- `sumexp__<s>__<g>`: Total read counts for subgenome `<s>` under condition `<g>`.
- `ratio__<s>__<g>`: HERs for subgenome `<s>` under condition `<g>`.
- `ratio_sd`: Standard deviation of HERs calculated from observed counts.


```{r, fc-hq-output}
head(x_output)
```


### Options

The number of iterations and number of samples drawn per iteration can be
adjusted using the `n_iters` and `iter_sampling` arguments, respectively.
For example:

```{r fc-hq-opts-iters, eval=FALSE}
x_output <- homeoroq(x, n_iters = 8, iter_sampling = 1e5)
```

Additionally, the `n_threads` option can be used to run iterations in parallel.

```{r fc-hq-opts-threds, eval=FALSE}
x_output <- homeoroq(x, n_threads = 8, n_iters = 8, iter_sampling = 1e5)
```




