# Installation

To install the stable version of the `r pkg("hespresso")` package
together with its required CRAN and Bioconductor dependencies,
start R and run:

```{r installation-package, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}

BiocManager::install("bitdessin/hespresso")
```

By default, HOBIT uses the maximum-likelihood estimation (MLE) implementation,
which does not require Stan [@ref_stan].
The optional Markov chain Monte Carlo (MCMC) implementation requires
the additional `r pkg("cmdstanr")` package and CmdStan,
which can be installed as follows:

```{r installation-cmdstanr, eval=FALSE}
install.packages("cmdstanr", repos = c("https://stan-dev.r-universe.dev", getOption("repos")))
cmdstanr::install_cmdstan()
```

::: {.alert .alert-info}

The development version of `r pkg("hespresso")` is available from
the [`devel` branch](https://github.com/bitdessin/hespresso/tree/devel) on GitHub.
For those interested in trying the latest features,
the development version can be installed by running:

```{r installation-devel-package, eval=FALSE}
BiocManager::install("bitdessin/hespresso", ref = "devel")
```

As the development version may include features that are still under evaluation,
the stable version is recommended for general use.

:::
