# Quick Start

The following example demonstrates the default HOBIT workflow,
from loading the input data to performing the statistical test.

```{r qs-hobit, results='hide', message=FALSE, warning=FALSE}
# Gene expression data
gexp <- read.table("data/c_flexuosa.0516.mini.txt.gz",
                   header = TRUE, sep = "\t", row.names = 1)

# Experimental groups
group <- c("wet", "wet", "wet", "dry", "dry", "dry")

# Mapping table for organizing gene expression into homeolog expression
mapping_table <- read.table("data/c_flexuosa.homeolog.txt.gz",
                            header = TRUE, sep = "\t")

# Prepare the input data
x <- newSeqCountData(gexp, group, mapping_table)

# Run the statistical test
x_output <- hobit(x)
```

```{r qs-hobit-output}
head(x_output)
```

By default, HOBIT internally normalizes library sizes and
performs statistical testing using maximum-likelihood estimation ([MLE](#hobit-mle)).
Therefore, the input must consist of raw, unnormalized integer RNA-seq read counts.
Do not provide transcripts per million (TPM),
fragments per kilobase of transcript per million mapped reads (FPKM),
counts per million (CPM), normalized counts, or log-transformed expression values.
